Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-10 fstats_pca_lewontin.tex
discussion

 However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.

>>>>

In many parts of the world, and particularly at more local scales, distinctions between populations begin to blur, and everyone could be considered  admixed to some degree \citep{pickrell2014}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as we do not need

. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
